### 2024-04-25 chap_RL/chap_RL.tex
SUBMISSION

Uncertainties about the future along with a large variety of \gls{IAMs} yield to an even larger variety of \gls{GHG} emissions reduction pathways \cite{nicolas2021robust}. For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track within the 1.5°C (if not, 2°C) increase of global temperature above pre-industrial levels by the end of the century. On the contrary, using their top-down model DICE, \citet{nordhaus2014question} state that immediate and drastic actions are not compulsory to meet the ambition of climate change mitigation.

To address this issue, several approaches have been used. Among them,  multi-stage stochastic programming is often put forward as a promising method. Stochastic programming formulates the problem as a mathematical program with probabilistic constraints or objective functions. These models explicitly consider the uncertainty by incorporating probability distributions for the uncertain parameters. The goal is to find an optimal decision that minimizes/maximizes the expected value of the objective function while satisfying the probabilistic constraints, modelled as a scenario tree.  At each stage of the problem, here the transition pathway of a whole-energy system, the model has the possibility of recourse, \ie to adapt the decisions made at earlier stages, in the response to unveiled uncertainties \cite{grossmann2016recent}. Using MARKAL model \cite{fishbone1981markal}, \citet{kanudia1998robust} assessed a multi-stage stochastic optimisation of the transition of Quebec between 1995 and 2035 accounting for high/low mitigation action plan and high/low growth scenarios. The authors found that hedging strategies, adapting with the future uncertainties, were outperforming the perfect foresight and deterministic optimisation of the different scenarios. However, stochastic programming is usually applied to limited number of uncertainties, \ie up to 10, and relies on probability distributions that are often difficult to define properly. Increasing the number of these uncertainties in stochastic programming increases exponentially the computational burden that limits the use of such a method in \gls{IAMs} \cite{nicolas2021robust}. Based on the approach of \citet{bertsimas2004price}, and similarly to \citet{Moret2017PhDThesis}, \citet{nicolas2021robust} rather opted for the robust optimisation of the global pathway up to 2200 given different temperature deviation targets, \ie 2 or 3°C by 2200 via the use of uncertainty budget, $\Gamma$, in the TIAM-World model \cite{loulou2005documentation}. Considering 9 climate parameters and their respective lower and upper bounds, the idea behind the uncertainty budget stems from the improbability of all parameters simultaneously reaching one of their two extreme values.  However, robust optimisation fails to provide a comprehensive hedging strategy, that can help guide near-term actions, like stochastic programming.

To keep the best of both worlds, \ie stochastic programming and robust optimisation, we decided to investigate the \gls{RL} approach, and its policy optimisation mechanism, to explore myopic transition pathways under uncertainties. Indeed, policymaking for transitioning a whole-energy system can be viewed as an iterative process of learning from policy implementation efforts, involving ongoing analysis of energy policy challenges and experimenting with various solutions \cite{howlett1995studying}. \gls{RL} exhibits two main advantages: its effectiveness to handle uncertainties and the model-free approach where an accurate representation of the real world is not needed to optimize the policy \cite{perera2021applications}.Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.  Then, the results of this policy optimisation point out strategies to follow, \ie \textit{sweet spots}, in the transitions under uncertainties as well as \textit{no-go zones} where the chances of succeeding the transition, \ie respecting the \ce{CO2}-budget, are very limited. Finally, these results are compared with references, \ie the perfect foresight and the myopic optimisation of the transition under the same uncertainties but without the trained \gls{RL}-agent that can support this transition thanks to its learned policy.

\section*{Contributions}
\label{sec:meth:contributions}
Applying the \gls{RL} approach to the optimization of the myopic transition pathway of a whole-energy system presents several novelties. First of all, as introduced in Section \ref{subsec:meth_RL_fundamentals}, when applied to energy systems, \gls{RL} is more dedicated either to smaller scale systems (\eg \gls{BEMS}, vehicles and energy devices) or to sector-specific, often the power sector, problems (\eg dispatch problems, energy markets and grid) \cite{perera2021applications}. In our case, the sector-coupling, the long-term goal at the end of a multiple-steps transition and the number of uncertain parameters make this application new for \gls{RL}.

Usually, applications of \gls{RL} focus more on the result of the learning process, the optimised policy, for optimising the control of system \cite{perera2021applications}. This thesis also investigates the learning episodes themselves to explore the field of possibilities to succeed the transition.

Then, applying to this optimisation environment, \ie EnergyScope myopic Pathway,  rather than a simulation environment, allows building a hierarchical multi-objective optimisation framework. In this agent, while the objective of the environment remains the minimisation of the total ``transition'' cost (on the concerned limited time window), the agent optimises its strategy to respect the \ce{CO2}-budget.

Finally, comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added-value brought by this approach in the exploration of myopic transition pathways.

\section{Definition of the actions, reward and states}
\label{sec:RL:act_states_rew}
As already introduced in Section \ref{subsec:meth_RL_algo}, the environment with which the \gls{RL}-agent interacts is the optimisation of the transition pathway whole-energy system on a specific time window, \eg 2020-2030 then 2025-2035 and so on, until 2040-2050 (see Figure \ref{fig:Schematics_RL}). In a nutshell, starting from the initial state of the environment (\ie the whole-energy system in 2020), the agent takes a set of actions that influence the environment, \ie that affects parameters of the Linear Programming in EnergyScope Pathway. Then, the window 2020-2030 is optimised via EnergyScope. Some of the outputs of this optimisation feed the agent with either the new state of the system or the reward, \ie telling the agent how good the actions were at the state he took it. Based on the new state and the reward, the agent takes another set of actions and the window 2025-2035 is optimised. This goes on until 2050.

@@

When the reward is not properly defined, the agent may optimize its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}. Even worse, it can lead to reward hacking (or reward tampering) where the agent exploits loopholes in the reward function to achieve higher rewards without actually performing the desired task \cite{amodei2016concrete}. On the contrary, a proper definition of the reward function increases the sample efficiency, \ie requiring less episode to converge to the optimal policy.  It also makes the policy more stable and able to withstand variations and uncertainties in the environment \cite{henderson2018deep}.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.65\textwidth]{Reward.pdf}
\caption{Reward function, $R$. Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded. If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues. Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost.}
\label{fig:Reward}
\end{figure}

The reward function is defined in three steps. First of all, taking a set of actions at a certain state might lead to an infeasible optimisation problem. In other words, as actions have a direct impact on some constraints of the problem, they might limit too much the feasible domain to the point where no solution can be found. For instance, the extreme case of aiming at carbon-neutrality, \ie $\mathrm{act}_{\mathrm{gwp}}=0$, and forbidding the use of the three aforementioned fossil fuels, \ie $\mathrm{act}\textsubscript{fossil gas}=\mathrm{act}\textsubscript{LFO}=\mathrm{act}\textsubscript{coal}=0$,  from the beginning of the transition makes the optimisation impossible to solve. In this case, the episode is prematurely ended and the reward is ``highly'' negative, -300. If the optimisation is feasible and the end of the transition, \ie 2050, is not reached, the cumulative emissions so far is evaluated. On the one hand, if these cumulative emissions exceed the \ce{CO2}-budget, $1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ (see Section \ref{sec:cs:CO2-budget}), the episode is also ended and a penalisation is given to the agent. This penalisation is proportional to the difference between the \ce{CO2}-budget and the actual cumulative emissions.  On the other hand, the episode continues with a zero reward if the \ce{CO2}-budget is not exceeded. Eventually, when reaching 2050, given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ambitious ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.  On top of that, the reward function includes a secondary objective: the cumulative transition cost. To make the agent sensitive to the cost-impact of its policy, we added the total transition cost in the reward function where the \emph{Trans. cost$_{\text{ref}}$} on Figure \ref{fig:Reward} is equal to $1.1\cdot10^3$\,b€. This value comes from the mean of the total transition costs obtained through the \gls{GSA} performed on the perfect foresight transition pathway optimisation (see Section \ref{subsec:atom_mol:results_uq_cost}). In this final form of the reward, one will notice that overshooting cumulative emissions are more penalising than an overshooting transition cost, \ie weight of 200 for the emissions versus 100 for the cost. The values of these weights are the results of a trial and error. This way, we observed that the agent first targeted the respect of the \ce{CO2}-budget and then, to a lesser scale, avoided reaching over-costly transitions.

@@

Consequently, after testing several state spaces and observing the convergence of the reward, we have converged to a four-dimension state space characterizing the energy system at the end of the optimised time window. The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2}-budget until 2050. Therefore, the cumulative emissions emitted so far in the current step of the transition is the first dimension of the states. Similarly, the cumulative cost of the transition so far constitutes the second dimension of the states to inform the agent about the cost-impact of its actions on the environment. Finally, to enrich the level of details, we have added two other dimensions representative of the key-to-the-transition indicators identified in the Renewable Energy Directive (RED) III of the European Commission \cite{REDIII}: the share of renewables in the primary energy mix and, the energy efficiency. The former is computed as the share of local renewables (\ie wind, solar, hydro and biomass) and imported renewable energy carriers (\ie biofuels and electrofuels) in the total consumption of primary energy. Electricity imported from abroad is not considered in the set of renewable energy carriers even if it can be assumed to be fully renewable by 2050. Finally, even though energy efficiency is usually defined as the ratio between the \gls{FEC} and the primary energy mix, we decided to define this efficiency with a focus on the \gls{EUD}, like in the rest of this thesis. Where electricity, heat and non-energy \gls{EUD} are expressed in terms of energy content, we needed to convert passenger and freight transports into their respective \gls{FEC} to integrate them in the ratio.

@@

Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the learning of the \gls{NN}, also called ``training''. For this, numerous episodes are played through the myopic optimisation of the transition pathway of Belgium. At the beginning of each episode, the agent starts with the Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}). Then, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window. This single draw will affect the parameters for all the subsequent time windows (see Figure \ref{fig:ranges_transition}).  Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward. This goes on until the end of the episode. For the new episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.

@@

The learning phase has been split into batches of 500 steps. %For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts. This makes sure replay buffer is full enough for useful updates. 
At the end of each batch, the up-to-date policy, \ie the \gls{NN}, is saved. This way, we can assess the progress in the learning process and its convergence (see Figure \ref{fig:reward_success}).  The mean reward increases rapidly at the beginning of the learning process before reaching a plateau where the optimisation of the policy becomes more marginal.  As this successes drive indirectly the agent's optimisation, it shows that the reward function (see Figure \ref{fig:Reward}) leads towards more and more successes.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Mean_reward.pdf}
\includegraphics[width=0.49\textwidth]{Success_rate.pdf}
\caption{Mean reward and success rate of the different learning batches. The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view. The evolution of the success rate also shows the reward function aims at more and more successful transitions.}
\label{fig:reward_success}
\end{figure}

During the learning process, the algorithm explores numerous transition pathways: 2037 successful transitions out of 10,751 attempts. Each pathway provides valuable insight into the possible alternatives --- the primary goal of reinforcement learning. As a side benefit, the collection of all explored pathways also identifies the intermediate milestones to reach and the range of actions that must be avoided or must be taken. Yet, the exploration during this learning process is not exhaustive. The trends provided below are therefore not proven. The randomness of the process and the number of explored transition pathways still give us high confidence.

There is a range in the reward where failures and successes overlap (see Figure \ref{fig:reward_status}). This area corresponds to either transitions that exceed the \ce{CO2}-budget in 2050 but are cheaper than the total transition cost of reference (see Section \ref{sec:RL:act_states_rew}) or successful transitions that are more expensive. Besides this overlap, we observe that successes account for the majority of the cases where the reward is positive. This is another indication that the reward function is appropriate in this exploration of successful transition pathways.

Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent. Beyond this point, through this learning process, the chances to succeed the transition were 38\%. In other words, near-term (2025-2030) actions are necessary to hope succeeding the transition.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Reward_status.pdf}
\includegraphics[width=0.49\textwidth]{Reward_status_2.pdf}
\caption{Reward distribution between successes and failures. Right hand side details at the end of which time-window the failure occurred. The ``tipping year'' is 2040 as failing the transition by 2040 represents 57\% of all the failures. Beyond this point, through this learning process, succeeding the transition represents 38\% of the episodes.}
\label{fig:reward_status}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.8\textwidth]{Cum_gwp_cost_2.pdf}
\caption{Exploration of the state space over the learning process: cumulative emissions and costs. The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case. }
\label{fig:Cum_gwp_cost}
\end{figure}

Given the reward function (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2}-budget. The skewness of the cumulative emissions and costs in 2050 are indications of this reward function (see Table \ref{tab:skewness_gwp_cost}). When succeeding the transitions, the cumulative emissions have a negative skewness: the agent successfully stayed within the budget and most of the cases where close to that budget (median at 0.9 Gt$_{\ce{CO2},\text{eq}}$). On the contrary, the cumulative cost of successful transitions has a positive skewness: the agent successfully reduces the cost of the system as a secondary objective with 30\% of the cases above the reference transition cost. This hierarchy of agent's objective is verified with the failures. When it failed the transition, the agent aimed at reducing the emissions (skewness of 0.61) before minimising the total transition cost (skewness of 0.24).

\begin{table}[htbp!]
\caption{Skewness of cumulative emissions and costs in 2050. Cumulative emissions are skewed to the left and to the right for the successes and failures, respectively. The skewness of the cumulative costs for successful transitions is higher compared to failures.On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.}
\label{tab:skewness_gwp_cost}
\centering
\begin{tabular}{l c c}
\toprule
\textbf{Status of episode}  & \textbf{Skewness of cumulative} & \textbf{Skewness of cumulative} \\
\textbf{in 2050}  & \textbf{emissions} & \textbf{costs} \\	
\midrule
Success & -0.52 & 0.50 \\
Failure & 0.61 & 0.24 \\
\bottomrule

@@

Finally, we observe that the majority of the successful transitions are cheaper than the reference transition cost, 1.1\,b€. Among the parameters impacting the most the total transition cost, we observe that success occurs when, on average,  the cost of purchasing fossil fuels is more increased than the one of electrofuels (see Table \ref{tab:param_RL}). In other words, to have higher chances to succeed a myopic transition, the key factor is to reduce the uncertainty on the cost of purchasing electrofuels. Given the skewness that is positive and negative for the electrofuels and the fossil fuels, respectively, these cases represent more than the majority of the successful cases. On top of this, total transition costs of successful episodes are lower due to lower industrial \gls{EUD} and interest rate. These favourable conditions combined with the right agent's actions led to transitions respecting the \ce{CO2}-budget.

@@

Besides the cumulative emissions and costs, the agent also observes the share of renewable energy carriers in the primary mix and the efficiency of the system The share of renewable energy carriers in the primary mix allows identifying intermediate milestones along successful transitions (see Figure \ref{fig:RE_in_mix_Efficiency}). From the initial state of 10\% in 2020, a boost of integration of renewables in the near-term is needed to hope for a successful transition. For the successful occurrences to exceed failures, this share increases to 54\% in 2025. Along the transitions, this increase goes with the import of electrofuels and the full deployment of local \gls{VRES}. In 2050, the success-failure threshold has been observed at 82\%. In the REF case of Chapter \ref{chap:atom_mol}, the share of renewables reached 86\% by 2050. However, by 2050, Figure \ref{fig:RE_in_mix_Efficiency} shows another threshold at lower share of renewables in the mix. This area corresponds to the possibility to install \gls{SMR}. As uranium is considered as a non-renewable resource \cite{rixhon2021terminology}, installing \gls{SMR} allows lowering the threshold to 60\% as in the SMR case of Chapter \ref{chap:atom_mol}. Besides these milestones to respect the \ce{CO2}-budget of the transition, one can also look at the other side of the thresholds. Below the near-term threshold of 60\%, this the ``no-go zone'' where succeeding the transition becomes unlikely, except if betting on the future installation of \gls{SMR}.

The efficiency, as defined in Section \ref{sec:RL:act_states_rew}, gives less valuable information to direct towards successful transition. Through the transition, besides the share of success increasing over the failures, these two scenarios indistinguishably spread over the whole range. Similarly than the emissions, we observe a bump at lower efficiencies by 2050 due to the installation of \gls{SMR}.

>>>>

Uncertainties about the future along with a large variety of \gls{IAMs} yield to an even larger variety of \gls{GHG} emissions reduction pathways \cite{nicolas2021robust}. For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track within the 1.5°C (if not, 2°C) increase of global temperature above pre-industrial levels by the end of the century. On the contrary, using his top-down model DICE, \citet{nordhaus2014question} states that immediate and drastic actions are not compulsory to meet the ambition of climate change mitigation.

To navigate through the myopic transitions not constrained by a prescribed \ce{CO2}-trajectory and investigate the efficiency of different policies, we apply the reinforcement learning approach detailed in Section \ref{sec:meth:RL} on the case study of Belgium. Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.  Then, the results of this policy optimisation point out strategies to follow, \ie \textit{sweet spots}, in the transitions under uncertainties as well as \textit{no-go zones} where the chances of succeeding the transition, \ie respecting the \ce{CO2}-budget, are very limited. Finally, these results are compared with references, \ie the perfect foresight and the myopic optimisation of the transition under the same uncertainties but without the trained \gls{RL}-agent that can support this transition thanks to its learned policy.

\section*{Contributions}
\label{sec:RL:contributions}
Applying the \gls{RL} approach to the optimization of the myopic transition pathway of a whole-energy system presents several novelties. First of all, as introduced in Section \ref{subsec:meth_RL_fundamentals}, when applied to energy systems, \gls{RL} is more dedicated either to smaller scale systems (\eg \gls{BEMS}, vehicles and energy devices) or to sector-specific, often the power sector, problems (\eg dispatch problems, energy markets and grid) \cite{perera2021applications}. In our case, the sector-coupling, the long-term goal at the end of a multiple-steps transition and the number of uncertain parameters make this application new for \gls{RL}.

Applying \gls{RL} to the optimisation environment of EnergyScope Pathway myopic,  rather than a simulation environment, allows building a hierarchical multi-objective optimisation framework: while the objective of the environment remains the minimisation of the total ``transition'' cost (on the concerned limited time window), the agent optimises its strategy to respect the \ce{CO2}-budget.

Comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added-value brought by this approach in the exploration of myopic transition pathways.

Usually, applications of \gls{RL} focus more on the result of the learning process, the optimised policy, for optimising the control of a system \cite{perera2021applications}. This thesis also investigates the learning episodes themselves to explore the field of possibilities to succeed the transition.

\section{Definition of the actions, reward and states}
\label{sec:RL:act_states_rew}
As already introduced in Section \ref{subsec:meth_RL_algo}, the environment with which the \gls{RL}-agent interacts is the optimisation of the transition pathway of a whole-energy system on a specific time window, \eg 2020-2030 then 2025-2035 and so on, until 2040-2050 (see Figure \ref{fig:Schematics_RL}). In a nutshell, starting from the initial state of the environment (\ie the whole-energy system in 2020), the agent takes a set of actions that influence the environment, \ie that affects parameters of the Linear Programming in EnergyScope Pathway. Then, the window 2020-2030 is optimised via EnergyScope. Some of the outputs of this optimisation feed the agent with either the new state of the system or the reward, \ie telling the agent how good the actions were at the state he took it. Based on the new state and the reward, the agent takes another set of actions and the window 2025-2035 is optimised. This goes on until 2050.

@@

When the reward is not properly defined, the agent may optimise its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}. Even worse, it can lead to reward hacking (or reward tampering) where the agent exploits loopholes in the reward function to achieve higher rewards without actually performing the desired task \cite{amodei2016concrete}. On the contrary, a proper definition of the reward function increases the sample efficiency, \ie requiring less episode to converge to the optimal policy.  It also makes the policy more stable and able to withstand variations and uncertainties in the environment \cite{henderson2018deep}.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.8\textwidth]{Reward.pdf}
\caption{Reward function, $R$. Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded. If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues. Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost.}
\label{fig:Reward}
\end{figure}

The reward function is defined in three steps. First of all, taking a set of actions at a certain state might lead to an infeasible optimisation problem. In other words, as actions have a direct impact on some constraints of the problem, they might limit too much the feasible domain to the point where no solution can be found. For instance, the extreme case of aiming at carbon-neutrality, \ie $\mathrm{act}_{\mathrm{gwp}}=0$, and forbidding the use of the three aforementioned fossil fuels, \ie $\mathrm{act}\textsubscript{fossil gas}=\mathrm{act}\textsubscript{LFO}=\mathrm{act}\textsubscript{coal}=0$,  from the beginning of the transition makes the optimisation impossible to solve. In this case, the episode is prematurely ended and the reward is ``highly'' negative, -300. If the optimisation is feasible and the end of the transition, \ie 2050, is not reached, the cumulative emissions so far are evaluated. On the one hand, if these cumulative emissions exceed the \ce{CO2}-budget, $1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ (see Section \ref{sec:cs:CO2-budget}), the episode is also ended and a penalisation is given to the agent. This penalisation is proportional to the difference between the \ce{CO2}-budget and the actual cumulative emissions.  On the other hand, the episode continues with a zero reward if the \ce{CO2}-budget is not exceeded. Eventually, when reaching 2050, given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.  On top of that, the reward function includes a secondary objective: the cumulative transition cost. To make the agent sensitive to the cost-impact of its policy, we added the total transition cost in the reward function where the \emph{Trans. cost$_{\text{ref}}$} on Figure \ref{fig:Reward} is equal to $1.1\cdot10^3$\,b€. This value comes from the mean of the total transition costs obtained through the \gls{GSA} performed on the perfect foresight transition pathway optimisation (see Section \ref{subsec:atom_mol:results_uq_cost}). In this final form of the reward, one will notice that overshooting cumulative emissions are more penalising than an overshooting transition cost, \ie weight of 200 for the emissions versus 100 for the cost. The values of these weights are the results of a trial and error to fine-tune the balance between more expensive successes and cheaper failures. This way, we observed that the agent first targeted the respect of the \ce{CO2}-budget and then, to a lesser scale, avoided reaching over-costly transitions.

@@

Consequently, after testing several state spaces and observing the convergence of the reward, we have converged to a four-dimension state space characterizing the energy system at the end of the optimised time window. The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2}-budget until 2050. Therefore, the cumulative emissions emitted so far up to the current step of the transition is the first dimension of the states. Similarly, the cumulative cost of the transition so far constitutes the second dimension of the states to inform the agent about the cost-impact of its actions on the environment. Finally, to enrich the level of details, we have added two other dimensions representative of the key-to-the-transition indicators identified in the Renewable Energy Directive (RED) III of the European Commission \cite{REDIII}: the share of renewables in the primary energy mix and, the energy efficiency. The former is computed as the share of local renewables (\ie wind, solar, hydro and biomass) and imported renewable energy carriers (\ie biofuels and electrofuels) in the total consumption of primary energy. Electricity imported from abroad is not considered in the set of renewable energy carriers even if it can be assumed to be fully renewable by 2050. Finally, even though energy efficiency is usually defined as the ratio between the \gls{FEC} and the primary energy mix, we decided to define this efficiency with a focus on the \gls{EUD}, like in the rest of this thesis. Where electricity, heat and non-energy \gls{EUD} are expressed in terms of energy content, we needed to convert passenger and freight transports into their respective \gls{FEC} to integrate them in the ratio. The information of efficiency fed back by the environment to the agent is the ratio between a ``hybrid'' \gls{EUD} and the consumption of primary energy resources.

@@

Before comparing the agent's optimised policy with references, the first step consists in assessing the learning of the \gls{NN}, also called ``training''. For this, numerous episodes are played through the myopic optimisation of the transition pathway of Belgium. At the beginning of each episode, the agent starts with the Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}). Then, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window. This single draw will affect the parameters for all the subsequent time windows (see Figure \ref{fig:ranges_transition}).  Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward. This goes on until the end of the episode. For the next episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), another sample of uncertain parameters is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.

@@

The learning phase has been split into batches of 500 steps. %For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts. This makes sure replay buffer is full enough for useful updates. 
At the end of each batch, the up-to-date policy, \ie the \gls{NN}, is saved. This way, we can assess the progress in the learning process and its convergence (see Figure \ref{fig:reward_success}).  The mean reward increases rapidly at the beginning of the learning process before reaching a plateau where the optimisation of the policy becomes more marginal.  As this successes drive indirectly the agent's optimisation, it shows that the reward function (see Figure \ref{fig:Reward}) leads towards more and more successes. However, given the wide range of uncertainty of some parameters and the agent's levers of action, this success rate stays limited at the end of the learning process. In other words, there are conditions where it is impossible for the agent to succeed the transition.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Mean_reward.pdf}
\includegraphics[width=0.49\textwidth]{Success_rate.pdf}
\caption{Mean reward and success rate of the different learning batches. The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view. The evolution of the success rate also shows that the reward function aims at more and more successful transitions.}
\label{fig:reward_success}
\end{figure}

\newpage
During the learning process, the algorithm explored numerous transition pathways: 2,037 successful transitions out of 10,751 attempts. 2\% of the total attempts led to infeasible optimisation of the initial time window. On top of that, each pathway provides valuable insight into the possible alternatives --- the primary goal of reinforcement learning. As a side benefit, the collection of all explored pathways also identifies the intermediate milestones to reach and the range of actions that must be avoided or must be taken. Yet, the exploration during this learning process is not exhaustive. The trends provided below are therefore not proven. The randomness of the process and the number of explored transition pathways still give high confidence in the conclusions drawn from these analyses.

There is a range in the reward where failures and successes overlap (see Figure \ref{fig:reward_status}). This area corresponds to either transitions that exceed the \ce{CO2}-budget in 2050 but are cheaper than the total transition cost of reference (see Section \ref{sec:RL:act_states_rew}) or successful transitions that are more expensive. Besides this overlap, we observe that successes account for the majority of the cases where the reward is positive. This is another indication that the reward function is appropriate in this exploration of successful transition pathways. More indirectly, this substantiates that the weights between emissions and cost defined in Section \ref{subsec:RL:act_states_rew:rew} could be used and adapted for another case study.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Reward_status.pdf}
\includegraphics[width=0.49\textwidth]{Reward_status_2.pdf}
\caption{Reward distribution between successes and failures. Right hand side details at the end of which time-window the failure occurred. The ``tipping year'' is 2040 as failing the transition by 2040 represents 57\% of all the failures. Beyond this point, through this learning process, succeeding the transition represents 38\% of the episodes. }
\label{fig:reward_status}
\end{figure}

\newpage
Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent. Beyond this point, through this learning process, the chances to succeed the transition were 38\%. In other words, near-term (2025-2030) actions are necessary to hope succeeding the transition.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.8\textwidth]{Cum_gwp_cost_3.pdf}
\caption{Exploration of the state space over the learning process: cumulative emissions and costs. The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case. }
\label{fig:Cum_gwp_cost}
\end{figure}

\newpage
Given the reward function (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2}-budget. The skewness of the cumulative emissions and costs in 2050 are indications of this reward function (see Table \ref{tab:skewness_gwp_cost}). When succeeding the transitions, the cumulative emissions have a negative skewness: the agent successfully stayed within the budget and most of the cases where close to that budget (median at 0.9 Gt$_{\ce{CO2},\text{eq}}$). On the contrary, the cumulative cost of successful transitions has a positive skewness: the agent successfully reduces the cost of the system as a secondary objective with 30\% of the cases above the reference transition cost. This hierarchy of agent's objective is verified with the failures. When it failed the transition, the agent aimed at reducing the emissions (skewness of 0.61) before minimising the total transition cost (skewness of 0.24).

\begin{table}[htbp!]
\caption{Skewness of cumulative emissions and costs in 2050. Cumulative emissions are skewed to the left and to the right for the successes and failures, respectively. The skewness of the cumulative costs for successful transitions is higher compared to failures. On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.}
\label{tab:skewness_gwp_cost}
\centering
\begin{tabular}{l c c}
\toprule
\textbf{Status of episode}  & \textbf{Skewness of cumulative} & \textbf{Skewness of cumulative} \\
\textbf{in 2050}  & \textbf{emissions} & \textbf{costs} \\	
\midrule
Success & -0.52 & 0.50 \\
Failure & 0.61 & 0.24 \\
\bottomrule

@@

\newpage
Finally, we observe that the majority of the successful transitions are cheaper than the reference transition cost, 1.1\,b€. Among the parameters impacting the most the total transition cost, we observe that success occurs when, on average,  the cost of purchasing fossil fuels is increased more than the one of electrofuels (see Table \ref{tab:param_RL}). In other words, to have higher chances to succeed a myopic transition, the key factor is to reduce the uncertainty on the cost of purchasing electrofuels or to increase the cost of fossil fuels. Given the skewness that is positive and negative for the electrofuels and the fossil fuels, respectively, these cases represent more than the majority of the successful cases. On top of this, total transition costs of successful episodes are lower due to lower industrial \gls{EUD} and interest rate. These favourable conditions combined with the right agent's actions led to transitions respecting the \ce{CO2}-budget.

@@

Besides the cumulative emissions and costs, the agent also observes the share of renewable energy carriers in the primary mix and the efficiency of the system The share of renewable energy carriers in the primary mix allows identifying intermediate milestones along successful transitions (see Figure \ref{fig:RE_in_mix_Efficiency}). From the initial state of 10\% in 2020, a boost of integration of renewables in the near-term is needed to hope for a successful transition. For the successful occurrences to exceed failures, this share increases to 54\% in 2025. Along the transitions, this increase goes with the import of electrofuels and the full deployment of local \gls{VRES}. In 2050, the threshold where occurences of success are more numerous than failures was at 82\% of renewable share. In the REF case of Chapter \ref{chap:atom_mol}, this share reached 86\% by 2050. However, by 2050, Figure \ref{fig:RE_in_mix_Efficiency} shows another ``bump'' at lower share of renewables in the mix. This area corresponds to the possibility to install \gls{SMR}. As uranium is considered as a non-renewable resource \cite{rixhon2021terminology}, installing \gls{SMR} allows lowering the threshold as in the SMR case of Chapter \ref{chap:atom_mol}. Besides these milestones to respect the \ce{CO2}-budget of the transition, one can also look at the other side of the thresholds. Below the near-term threshold of $\sim$60\%, this is the ``no-go zone'' where succeeding the transition becomes unlikely, except if betting on the future installation of \gls{SMR}.

The efficiency, as defined in Section \ref{sec:RL:act_states_rew}, gives less valuable information towards successful transitions. Through the transition, besides the share of success increasing over the failures, the distributions of success and failure indistinguishably spread over the whole range. Similarly than the emissions, we observe a bump at lower efficiencies by 2050 due to the installation of \gls{SMR}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.8\textwidth]{RE_in_mix_Efficiency.pdf}
\caption{Exploration of the state space over the learning process: share of renewable energy carriers in the primary energy mix and efficiency. Integration of local \gls{VRES} at early stages then massive import of electrofuels later are needed to secure successful transitions. Below a near-term threshold ($\sim$60\%), the chances of success are limited, \ie no-go zones. Efficiency is a less valuable information for the agent to succeed transitions as failures and successes indistinguishably spread over the whole range.}
\label{fig:RE_in_mix_Efficiency}
\end{figure}
